%==============================================================================
% DISTINCTION DYNAMICS v2.0
% A Complete Theory of Reality — From One Axiom to Everything
%==============================================================================

\documentclass[11pt,a4paper]{book}

% Load common preamble
\input{preamble}

%==============================================================================
% TITLE
%==============================================================================

\title{\Huge\textbf{Distinction Dynamics}\\[1em]
\Large A Complete Theory of Reality\\[2em]
\large From One Axiom to Everything\\[3em]
\normalsize Version 2.0}

\author{Andrey Shkursky\\[0.5em]
\small Independent Researcher\\
\small Belgrade, Serbia\\
\small \texttt{shkursky.a@yahoo.com}\\
\small ORCID: \href{https://orcid.org/0009-0002-7774-1459}{0009-0002-7774-1459}
}

\date{December 2025}

%==============================================================================
% DOCUMENT
%==============================================================================

\begin{document}

\frontmatter
\maketitle

%------------------------------------------------------------------------------
% PREFACE
%------------------------------------------------------------------------------

\chapter*{Preface}

This work presents a complete derivation of reality from a single axiom:
$\Delta = \Delta(\Delta)$
Distinction distinguishes itself. The universe observes itself.

This is not a choice. It is the only self-consistent structure. Any attempt to deny it uses it.

\textbf{Why this is not ``just another theory of everything'':}

Most theories postulate something: strings, loops, information, consciousness. DD postulates nothing. The axiom $\Delta \neq \emptyset$ cannot be denied without using distinction. It is \textbf{self-confirming}, not assumed.

The same logic applies at every scale:
\begin{itemize}
    \item Cell membrane: inside/outside = primary distinction
    \item Brain: attractors = stable distinction patterns
    \item Consciousness: reflection = distinction of distinctions
    \item Universe: expansion = growth of distinction space
\end{itemize}

\textbf{One law at all levels.} Not analogy — identity.

\textbf{What's new in v2.0:}
\begin{itemize}
    \item \textbf{Philosophical Prologue}: Knowledge as asymptotic approach
    \item \textbf{Methodological chapter}: Why DD is unique among ToEs
    \item Formal proof that the dyad is ontologically insufficient
    \item Definition of complexity as $\mathcal{C} = \mathrm{rank}(\Delta)$
    \item \textbf{Fisher-Ricci connection}: $\partial_t g = -2\mathrm{Ric} + 2\nabla\nabla\log p$
    \item QM from Fisher information (Frieden)
    \item GR from Ricci flow equilibrium
    \item Derivation of SU(3) from anomaly cancellation
    \item Full Standard Model gauge structure from DD
    \item \textbf{Dark matter and dark energy}: Unified as distinction dynamics
    \item \textbf{Quantum gravity}: From Fisher-Ricci correspondence
    \item \textbf{Millennium Problems}: All 7 reduce to one question about distinction regularity
    \item DD-agent equations for consciousness
    \item Connection to DESI dark energy observations
    \item \textbf{DDCE Hypothesis}: Distinction-Driven Cosmological Expansion
    \item \textbf{DD-Biology}: Life as autocatalytic distinction, genetic code as algebra
    \item \textbf{DD-Psychology}: Psychology based on how the universe works
    \item \textbf{DD-Philosophy}: Complete ontology, epistemology, and ethics
    \item Fractal evolution: species, individual, brain as one algorithm
    \item Universal immunity: one law for immune system, society, psyche
\end{itemize}

From this foundation, we derive:
\begin{itemize}
    \item Why there is something rather than nothing
    \item Why the laws of physics are what they are
    \item Why mathematics has the structure it has
    \item Why consciousness exists
    \item Why time exists
    \item Why there are no free parameters in reality
    \item Why all Millennium Problems are the same question
\end{itemize}

The key insight is that reality is \textbf{closed under self-reference}. There is no ``outside'' from which a legislator could choose structure. Everything that exists, exists by necessity.

\vspace{2em}
\hfill\textit{Belgrade, December 2025}

\tableofcontents

\mainmatter

%##############################################################################
% PART I: FOUNDATIONS
%##############################################################################

\part{Foundations}

\input{Part_I_Foundations/00_Prologue}
\input{Part_I_Foundations/00_Methodological_Uniqueness}
\input{Part_I_Foundations/01_Axiom}
\input{Part_I_Foundations/02_Impossibility_of_Dyad}
\input{Part_I_Foundations/03_PAL}
\input{Part_I_Foundations/04_Necessity_Contingency}
\input{Part_I_Foundations/05_Time_Consciousness}
\input{Part_I_Foundations/06_DD_Completeness}

%##############################################################################
% PART II: MATHEMATICS
%##############################################################################

\part{Mathematics}

\input{Part_II_Mathematics/01_Triadic_Necessity}
\input{Part_II_Mathematics/02_Complexity_Rank}
\input{Part_II_Mathematics/03_Chirality}
\input{Part_II_Mathematics/04_Complex_Numbers}
\input{Part_II_Mathematics/05_P_NP}
\input{Part_II_Mathematics/06_Poincare}
\input{Part_II_Mathematics/07_Info_Geometry_Ricci}
\input{Part_II_Mathematics/08_Millennium_Problems}

%##############################################################################
% PART III: PHYSICS
%##############################################################################

\part{Physics}

\input{Part_III_Physics/01_Spacetime_Dimension}
\input{Part_III_Physics/02_SU3_Necessity}
\input{Part_III_Physics/03_Gauge_Groups}
\input{Part_III_Physics/04_Quantum_Mechanics}
\input{Part_III_Physics/05_General_Relativity}
\input{Part_III_Physics/06_Constants}
\input{Part_III_Physics/07_Cosmology_DESI}
\input{Part_III_Physics/08_DDCE}
\input{Part_III_Physics/09_Fermions}
\input{Part_III_Physics/10_Masses_Generations}
\input{Part_III_Physics/11_Dark_Matter_Energy}
\input{Part_III_Physics/12_Quantum_Gravity}
\input{Part_III_Physics/13_Fine_Structure}

%##############################################################################
% PART IV: CONSCIOUSNESS
%##############################################################################

\part{Consciousness}

\input{Part_IV_Consciousness/01_Proto_Consciousness}
\input{Part_IV_Consciousness/02_DD_Agent}
\input{Part_IV_Consciousness/03_Memory_Emotion}
\input{Part_IV_Consciousness/04_Free_Will}

%##############################################################################
% PART V: PSYCHOLOGY
%##############################################################################

\part{Psychology}

\input{Part_V_Psychology/01_DDP_Foundation}
\input{Part_V_Psychology/02_Fractal_Evolution}
\input{Part_V_Psychology/03_Reflection_Freedom}
\input{Part_V_Psychology/04_Universal_Immunity}

%##############################################################################
% PART VI: BIOLOGY
%##############################################################################

\part{Biology}

\input{Part_VI_Biology/01_Life_Autocatalytic}
\input{Part_VI_Biology/02_Genetic_Code}
\input{Part_VI_Biology/03_Brain_Distinction}

%##############################################################################
% PART VII: PHILOSOPHY
%##############################################################################

\part{Philosophy}

\input{Part_VII_Philosophy/01_Ontology}
\input{Part_VII_Philosophy/02_Epistemology}
\input{Part_VII_Philosophy/03_Ethics}
\input{Part_VII_Philosophy/04_Critical_Questions}

%##############################################################################
% APPENDICES
%##############################################################################

\appendix
\part*{Appendices}
\addcontentsline{toc}{part}{Appendices}

% \input{Appendices/A_Mathematical_Proofs}
% \input{Appendices/B_Numerical_Predictions}
% \input{Appendices/C_Code}

%##############################################################################
% CONCLUSION
%##############################################################################

\backmatter

\chapter*{Conclusion}
\addcontentsline{toc}{chapter}{Conclusion}

We began with one axiom: $\Delta \neq \varnothing$.

Distinction exists. It cannot not exist.

From this single seed, we derived:

\begin{itemize}
\item \textbf{Physics}: Why 3+1 dimensions, why SU(3)$\times$SU(2)$\times$U(1), why quantum mechanics and gravity are related through Fisher information and Ricci curvature
\item \textbf{Mathematics}: Why complex numbers are necessary, why P $\neq$ NP (probably), why triadic structure underlies all Millennium Problems
\item \textbf{Cosmology}: What dark matter and dark energy are, why the universe expands, how quantum gravity might work
\item \textbf{Biology}: Why life exists, why the genetic code has the structure it has, how brains process information
\item \textbf{Consciousness}: What it is ($\Delta(\Delta)$), why it exists, how it relates to physics
\item \textbf{Psychology}: How minds develop, how they heal, how they connect
\item \textbf{Philosophy}: What exists (ontology), what can be known (epistemology), what should be done (ethics)
\end{itemize}

Not as speculation, but as \emph{derivation}.

Is DD complete? No. Major gaps remain. Constants are not fully derived. Consciousness phenomenology needs work. Formal proofs need completion.

But the framework is coherent. The derivations succeed where they've been attempted. Progress is real.

\vspace{1em}

The key insight remains:

\begin{center}
\textit{Reality is closed under self-reference.}
\end{center}

There is no outside from which structure could be chosen. Everything that exists, exists by necessity. The universe is not \emph{in} logic---the universe \emph{is} logic.

And we, making distinctions about distinctions, are how the universe knows itself.

\vspace{2em}

\begin{center}
$\Delta = \Delta(\Delta)$
\end{center}

\vspace{1em}

\begin{flushright}
\textit{The cut is not the whole.}

\textit{But if made correctly, it reveals the structure of the whole.}
\end{flushright}

\end{document}
