%==============================================================================
% Chapter: DD-Completeness
% Part I: Foundations
%==============================================================================

\chapter{DD-Completeness}\label{ch:dd-complete}

\epigraph{Zero free parameters.}{---}

\section{Definition}

\begin{definition}[DD-Complete]\label{def:dd-complete}
A theory $\mathcal{T}$ is \textbf{DD-complete} if:
\begin{enumerate}[label=(\roman*)]
    \item All structures in $\mathcal{T}$ follow from $\Self$
    \item $\mathcal{T}$ contains no free parameters
    \item $\mathcal{T}$ contains no choice points
\end{enumerate}
\end{definition}

\section{Reality is DD-Complete}

\begin{theorem}[DD-Completeness of Reality]\label{thm:reality-complete}
Physical reality is DD-complete.
\end{theorem}

\begin{proof}
\begin{enumerate}
    \item Reality exists (Corollary~\ref{cor:exists}).
    \item Everything in reality is necessary (Theorem~\ref{thm:no-contingency}).
    \item Necessary structures follow from $\Self$ (the unique self-grounding axiom).
    \item Free parameters require selection among values.
    \item Selection requires either:
    \begin{enumerate}[label=(\alph*)]
        \item A legislator, or
        \item Determination by $\Delta$
    \end{enumerate}
    \item (a) is impossible (Theorem~\ref{thm:no-legislator}).
    \item Therefore, (b): all values are determined by $\Delta$.
    \item Determined values are not free.
    \item Therefore, no free parameters exist.
    \item Choice points require an external chooser or determination by $\Delta$.
    \item No external chooser exists. $\Delta$ determines.
    \item No genuine choice points exist.
    \item Reality is DD-complete. \qedhere
\end{enumerate}
\end{proof}

\section{Consequences}

\begin{corollary}[Computability of Constants]
All fundamental constants are derivable from $\Self$.
\end{corollary}

\begin{remark}
The 19+ parameters of the Standard Model (quark masses, mixing angles, coupling constants, etc.) are not free. They follow from the unique self-consistent realization of $\Delta$-structure in physics.
\end{remark}

\begin{corollary}[No Fine-Tuning]
The apparent fine-tuning of constants is not tuning---it is necessity.
\end{corollary}

\begin{corollary}[No Multiverse]
Multiple universes with different constants are impossible.
\end{corollary}

\begin{proof}
Different constants = different realizations of ``universe.'' By Theorem~\ref{thm:uniqueness}, at most one realization exists.
\end{proof}

\begin{corollary}[Predictive Power]
DD is falsifiable: if any constant is shown to be arbitrary (not derivable from $\Delta$), DD is false.
\end{corollary}

\section{Complexity and DD-Completeness}

\vnew{Connection to Chapter~\ref{ch:dyad}.}

\begin{theorem}[Complexity Bound]
The complexity $\Complexity = \rank(\Delta)$ of any DD-complete system is uniquely determined.
\end{theorem}

\begin{proof}
\begin{enumerate}
    \item $\Complexity$ is a structural property of $\Delta$.
    \item By DD-completeness, all structural properties are determined by $\Self$.
    \item Therefore, $\Complexity$ is determined, not free.
    \item The minimal $\Complexity = 2$ (triad) is necessary for self-sufficiency.
    \item Higher $\Complexity$ follows from elaboration of the triadic structure. \qedhere
\end{enumerate}
\end{proof}

\section{Summary of Part I}

\subsection{Logical Dependencies}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Result} & \textbf{Depends On} & \textbf{Status} \\
\midrule
Axiom: $\Self$ & --- & Undeniable \\
Lemma: Assertion = distinction & --- & Proved \\
Lemma: $\Delta$ is primitive & --- & Proved \\
Lemma: Unity of $\Delta$ & Structure lemma & Proved \\
Theorem: Self-application & Unity, Existence def & Proved \\
Theorem: Undeniability & Assertion, Unity & Proved \\
Corollary: $\Delta \neq \emptyset$ & Axiom & Proved \\
Theorem: Closure & Unity & Proved \\
\vnew{Theorem: Dyad insufficient} & \vnew{New in v2} & \vnew{Proved} \\
\vnew{Theorem: Triad minimal} & \vnew{New in v2} & \vnew{Proved} \\
Theorem: No legislator & Closure or Unity & Proved \\
Theorem: No randomness & Unity, Axiom & Proved \\
Theorem: No contingency & Unity, Axiom & Proved \\
Theorem: Time exists & Axiom (recursion) & Proved \\
Theorem: Proto-consciousness & Axiom & Proved \\
Theorem: DD-completeness & No contingency, PAL & Proved \\
\bottomrule
\end{tabular}
\end{center}

\subsection{The Closed Loop}

\begin{center}
\fbox{\parbox{0.85\textwidth}{\centering
\textbf{The universe observes itself.}\\[0.5em]
$\Self$\\[1em]
\textbf{Closure}: Nothing outside $\Delta$.\\
\textbf{PAL}: No legislator.\\
\textbf{Triadic Necessity}: Dyad insufficient, triad minimal.\\
\textbf{Necessity}: No contingency.\\
\textbf{Self-Determination}: No randomness.\\
\textbf{Completeness}: No free parameters.\\[0.5em]
Time and proto-consciousness are intrinsic.\\[0.5em]
Everything else follows.
}}
\end{center}

\subsection{What Remains}

Parts II--IV derive specific structures from this foundation:
\begin{itemize}
    \item \textbf{Mathematics}: Triadic necessity, chirality, complex numbers, P$\neq$NP, Poincar\'e.
    \item \textbf{Physics}: Spacetime (3+1), gauge groups, quantum mechanics, general relativity.
    \item \textbf{Consciousness}: DD-agent, memory, emotion, free will.
\end{itemize}

Each derivation shows how the structure follows necessarily from $\Self$.
