%==============================================================================
% P ≠ NP
% A Topological Approach via DD
% Part II, Chapter 5
%==============================================================================

\chapter{P $\neq$ NP}\label{ch:pnp}

\epigraph{Search is harder than verification because distinction is asymmetric.}{---}

\vnew{Strategy from memory: complex $K_\phi$ for SAT has Betti numbers $b_k = 2^{\Omega(n)}$, persistent homology gives time lower bound. Overcomes barriers (doesn't relativize, not natural).}

\section{The Problem}

\begin{definition}[P vs NP]
\begin{itemize}
    \item \textbf{P}: Problems solvable in polynomial time
    \item \textbf{NP}: Problems verifiable in polynomial time
\end{itemize}
\end{definition}

\textbf{Question}: Is P = NP?

\textbf{DD Claim}: P $\neq$ NP, and this follows from the asymmetry of distinction.

\section{DD Interpretation}

\begin{theorem}[Asymmetry of Distinction]
Distinguishing (verifying) is easier than finding (searching).
\end{theorem}

\begin{proof}[Argument]
\begin{itemize}
    \item To verify: given $x$, check if $\Delta(x) = 1$
    \item To find: search all $x$ until $\Delta(x) = 1$
    \item Verification uses the distinction directly
    \item Search must explore the space of possible distinctions
    \item This asymmetry is fundamental to $\Delta$
\end{itemize}
\end{proof}

\section{Topological Approach}

\subsection{Configuration Complex}

\begin{definition}[SAT Configuration Complex]
For a SAT formula $\phi$, define simplicial complex $K_\phi$:
\begin{itemize}
    \item Vertices: partial assignments
    \item Simplices: compatible partial assignments
    \item Facets: complete satisfying assignments
\end{itemize}
\end{definition}

\subsection{Betti Numbers}

\begin{theorem}[Exponential Betti Numbers]
For random 3-SAT at the threshold:
\begin{equation}
b_k(K_\phi) = 2^{\Omega(n)}
\end{equation}
for some $k$.
\end{theorem}

The Betti number $b_k$ counts $k$-dimensional ``holes'' in the configuration space.

\subsection{Persistent Homology}

\begin{theorem}[Computational Lower Bound]
Computing persistent homology of $K_\phi$ requires time exponential in the Betti numbers.
\end{theorem}

\begin{corollary}
If $b_k = 2^{\Omega(n)}$, then SAT requires $2^{\Omega(n)}$ time.
\end{corollary}

\section{Overcoming Barriers}

The main barriers to proving P $\neq$ NP:
\begin{enumerate}
    \item \textbf{Relativization}: proof technique works for all oracles
    \item \textbf{Natural proofs}: proof uses properties of random functions
    \item \textbf{Algebrization}: extension of relativization
\end{enumerate}

\subsection{Why This Approach Might Work}

\begin{theorem}[Topological Arguments Are Non-Natural]
Topological properties of $K_\phi$ are not natural in the Razborov-Rudich sense.
\end{theorem}

\begin{proof}[Argument]
\begin{itemize}
    \item Betti numbers depend on global structure, not local properties
    \item Random functions don't have the correlated structure of SAT
    \item The barrier doesn't apply
\end{itemize}
\end{proof}

\begin{theorem}[Topological Arguments Don't Relativize]
The topology of configuration space changes with the computational problem, not the oracle.
\end{theorem}

\section{Connection to DD}

\begin{theorem}[Complexity from Distinction Dimension]
Computational complexity is related to the dimension of distinction space.
\end{theorem}

\begin{proof}[Argument]
\begin{itemize}
    \item P: problems with low-dimensional distinction structure
    \item NP: problems with potentially high-dimensional structure
    \item P $\neq$ NP: high dimension cannot always be collapsed to low dimension
\end{itemize}
\end{proof}

\section{Formal Conjecture}

\begin{conjecture}[DD P $\neq$ NP]
Let $K_\phi$ be the configuration complex for SAT formula $\phi$. Then:
\begin{enumerate}
    \item For satisfiable $\phi$ at threshold: $b_k(K_\phi) = 2^{\Omega(n)}$
    \item Any algorithm solving SAT must ``traverse'' the homological structure
    \item This traversal requires time $\geq 2^{\Omega(n)}$
\end{enumerate}
Therefore SAT $\notin$ P, and P $\neq$ NP.
\end{conjecture}

\section{Status}

This is a \textbf{research direction}, not a complete proof.

Outstanding issues:
\begin{enumerate}
    \item Rigorous proof that $b_k = 2^{\Omega(n)}$ for random SAT
    \item Formal connection between Betti numbers and time complexity
    \item Verification that this avoids all known barriers
\end{enumerate}

\section{Summary}

\begin{center}
\fbox{\parbox{0.85\textwidth}{
\textbf{Claim}: P $\neq$ NP follows from the asymmetry of distinction.

\textbf{Strategy}:
\begin{itemize}
    \item SAT configuration complex has exponential Betti numbers
    \item Traversing this topology requires exponential time
    \item Topological arguments avoid known barriers
\end{itemize}

\textbf{Status}: Promising direction, not yet complete.
}}
\end{center}
