%==============================================================================
% PHYSICAL CONSTANTS FROM DD
% Constraints and Relations
% Part III, Chapter 6
%==============================================================================

\chapter{Physical Constants}\label{ch:constants}

\epigraph{The constants are not chosen. They are computed — or they are impossible.}{---}

\vnew{This chapter explores what DD can and cannot say about constants.}

\section{The Problem of Constants}

The Standard Model has $\sim 19$ free parameters:
\begin{itemize}
    \item 3 gauge couplings: $g_1, g_2, g_3$
    \item 9 Yukawa couplings (fermion masses)
    \item 4 CKM parameters
    \item Higgs parameters: $\mu, \lambda$
    \item $\theta_{QCD}$
\end{itemize}

Plus gravitational constant $G$ and cosmological constant $\Lambda$.

\textbf{Can DD derive any of these?}

\section{What DD Claims}

From Chapter~\ref{ch:dd-complete} (DD-Completeness):

\begin{theorem}[Zero Free Parameters]
In the final theory, all parameters are fixed by self-consistency. There are no genuinely free choices.
\end{theorem}

However, DD in its current form does not compute specific values.

What DD \textbf{does} provide:
\begin{enumerate}
    \item Relations between constants
    \item Constraints from consistency
    \item Qualitative predictions
\end{enumerate}

\section{Coupling Constant Unification}

\subsection{Running Couplings}

The gauge couplings run with energy:
\begin{equation}
\frac{1}{\alpha_i(\mu)} = \frac{1}{\alpha_i(M_Z)} - \frac{b_i}{2\pi} \ln\frac{\mu}{M_Z}
\end{equation}

\subsection{DD Prediction}

\begin{theorem}[Coupling Relation]
At some scale $M_U$, all gauge couplings should unify:
\begin{equation}
\alpha_1(M_U) = \alpha_2(M_U) = \alpha_3(M_U)
\end{equation}
\end{theorem}

\begin{proof}[Argument from DD]
\begin{enumerate}
    \item At high energy, all distinctions are equivalent
    \item Distinct gauge groups emerge from one structure
    \item Therefore, couplings should match at unification scale
\end{enumerate}
\end{proof}

\textbf{Observation}: In the SM, couplings nearly unify at $M_U \sim 10^{16}$ GeV.

With supersymmetry, unification is exact.

\section{The Fine Structure Constant}

\subsection{The Value}

\begin{equation}
\alpha = \frac{e^2}{4\pi\epsilon_0 \hbar c} \approx \frac{1}{137.036}
\end{equation}

\subsection{DD Approach: Status BOUND}

The fine structure constant $\alpha$ has status \textbf{BOUND}:
\begin{itemize}
    \item \textbf{Value}: Constrained to critical window (not arbitrary)
    \item \textbf{Formula}: Specific derivation remains CONJ
\end{itemize}

\subsubsection{Criticality Constraint}

\begin{theorem}[Critical Window for $\alpha$]
$\alpha$ must be $O(1/100)$ for stable gauge coupling:
\begin{itemize}
    \item $\alpha \ll 1$: Required for perturbative QED (convergent expansion)
    \item $\alpha \gtrsim 1$: Would cause strong QED ($e^+e^-$ pair creation, no stable atoms)
    \item The window $\alpha \in (1/200, 1/50)$ is the stable regime
\end{itemize}
\end{theorem}

This constraint is \textbf{FORCED} by criticality (0 < $\Phi$ < $\infty$).

\subsubsection{Numerical Patterns (CONJ)}

Multiple numerical observations converge on $\alpha^{-1} \approx 137$:

\textbf{Pattern 1: Geometric series}
\begin{equation}
\alpha^{-1} \approx 4\pi^3 + \pi^2 + \pi = 137.04...
\end{equation}
Interpretation: Powers of $\pi$ suggesting geometric volume structure.

\textbf{Pattern 2: Gauge structure}
\begin{equation}
\alpha^{-1} = 11^2 + 4^2 = 121 + 16 = 137
\end{equation}
where:
\begin{itemize}
    \item $11 = 3 + 8$ (dim SU(2) + dim SU(3))
    \item $4 = 2^2$ (dyadic structure)
\end{itemize}

\textbf{Pattern 3: Running to unification}
\begin{equation}
\alpha(M_Z) \to \alpha(M_{GUT}) \to \alpha_{unified}
\end{equation}
The low-energy value $\alpha \approx 1/137$ may be fixed by unification boundary conditions.

\subsubsection{Status Assessment}

\begin{center}
\begin{tabular}{lcl}
\hline
Claim & Status & Confidence \\
\hline
$\alpha \in (1/200, 1/50)$ & BOUND & 90\% \\
$\alpha^{-1} = 4\pi^3 + \pi^2 + \pi$ & CONJ & 30\% \\
$\alpha^{-1} = 11^2 + 4^2$ & CONJ & 40\% \\
Unification determines $\alpha$ & UNTRACED & 60\% \\
\hline
\end{tabular}
\end{center}

\textbf{Verdict}: The \textit{order of magnitude} of $\alpha$ is constrained by criticality. The \textit{exact value} remains an open problem. Numerical patterns may be coincidence or deep structure.

\section{Mass Ratios}

\subsection{Proton-Electron Mass Ratio: $m_p/m_e = 6\pi^5$}

The proton-to-electron mass ratio has a remarkable structural formula:
\begin{equation}
\boxed{\frac{m_p}{m_e} = 6\pi^5 = N_c \times \text{Vol}(S^3) \times \text{Vol}(S^5) = 3 \times 2\pi^2 \times \pi^3}
\end{equation}

\begin{center}
\begin{tabular}{lcc}
\hline
& Theoretical & Experimental \\
\hline
$m_p/m_e$ & 1836.118 & 1836.153 \\
\hline
Agreement & \multicolumn{2}{c}{99.998\%} \\
\hline
\end{tabular}
\end{center}

\textbf{Physical interpretation}:
\begin{itemize}
    \item $N_c = 3$: Color singlet (triad)
    \item $\text{Vol}(S^5) = \pi^3$: Three quarks in $\mathbb{R}^6$ relative space
    \item $\text{Vol}(S^3) = 2\pi^2$: Isospin SU(2) structure
\end{itemize}

\textbf{Status}: Mathematics PROVEN; axiom $m \propto \Omega$ is CONJ (85\% confidence).

\textbf{Full derivation}: See Chapter~\ref{chap:mass_ratio}.

\subsection{Fermion Mass Hierarchy}

\begin{equation}
\frac{m_t}{m_e} \sim 3.4 \times 10^5
\end{equation}

This enormous hierarchy is unexplained in SM.

\subsection{DD Perspective}

\begin{theorem}[Mass from Distinction Strength]
Fermion mass $\propto$ coupling to Higgs $\propto$ distinction strength with vacuum.
\end{theorem}

Different generations have different ``positions'' in triadic structure:
\begin{itemize}
    \item 3rd generation: closest to triadic center (strongest coupling)
    \item 1st generation: furthest from center (weakest coupling)
\end{itemize}

Mass ratios follow exponential scaling with spectral eigenvalues. See Chapter~\ref{ch:masses}.

\section{Planck Units}

\subsection{Natural Units}

\begin{align}
l_P &= \sqrt{\frac{\hbar G}{c^3}} \approx 1.6 \times 10^{-35} \text{ m} \\
t_P &= \sqrt{\frac{\hbar G}{c^5}} \approx 5.4 \times 10^{-44} \text{ s} \\
m_P &= \sqrt{\frac{\hbar c}{G}} \approx 2.2 \times 10^{-8} \text{ kg}
\end{align}

\subsection{DD Interpretation}

\begin{theorem}[Planck Scale as Distinction Limit]
The Planck scale is where distinction itself becomes quantum:
\begin{equation}
\Delta x \cdot \Delta p \geq \frac{\hbar}{2}
\end{equation}
At $l_P$, the uncertainty in distinction equals the distinction itself.
\end{theorem}

Therefore: Planck scale = \textbf{resolution limit of distinction}.

\section{The Cosmological Constant Problem}

\subsection{The Puzzle}

Observed: $\Lambda \sim 10^{-122} M_P^4$

QFT prediction: $\Lambda \sim M_P^4$

Discrepancy: 122 orders of magnitude!

\subsection{DD Resolution (DDCE)}

From Chapter~\ref{ch:ddce}:

\begin{equation}
\Lambda_{\text{eff}} = k(\Delta + F + M)
\end{equation}

The cosmological ``constant'' is \textbf{not constant} — it's determined by the current state of distinction complexity.

\begin{theorem}[Dynamic $\Lambda$]
The small observed $\Lambda$ reflects the current cosmic epoch's distinction structure, not a fundamental constant.
\end{theorem}

This connects to DESI observations of evolving dark energy.

\section{The Hierarchy Problem}

\subsection{The Puzzle}

Why is the Higgs mass ($\sim 125$ GeV) so much smaller than the Planck scale ($\sim 10^{19}$ GeV)?

\subsection{DD Perspective}

\begin{theorem}[Hierarchy from Information Levels]
The hierarchy reflects the ratio between:
\begin{itemize}
    \item Fundamental distinction scale (Planck)
    \item Electroweak distinction scale (Higgs)
\end{itemize}
\end{theorem}

The Higgs scale is where \textbf{matter distinctions crystallize}.

This is lower than the \textbf{spacetime distinction scale} (Planck).

Full derivation remains open.

\section{What DD Predicts}

\subsection{Qualitative}

\begin{enumerate}
    \item Coupling unification at high energy
    \item Mass hierarchy correlates with generation number
    \item $\Lambda$ evolves with cosmic time
    \item Planck scale = distinction resolution limit
\end{enumerate}

\subsection{Testable Relations}

\begin{theorem}[DDCE Prediction]
\begin{equation}
w(z) \neq -1
\end{equation}
The dark energy equation of state evolves with redshift.
\end{theorem}

This is being tested by DESI, Euclid, and future surveys.

\section{What DD Does Not (Yet) Derive}

\begin{itemize}
    \item $\alpha = 1/137.036...$
    \item $m_e/m_p = 1/1836...$
    \item Specific Yukawa couplings
    \item CKM matrix elements
    \item $\theta_{QCD} \approx 0$
\end{itemize}

These may require:
\begin{itemize}
    \item Deeper understanding of triadic number theory
    \item Explicit computation on Fisher-Ricci manifold
    \item Connection to string/M-theory landscape
\end{itemize}

\section{Summary}

\begin{center}
\fbox{\parbox{0.9\textwidth}{
\textbf{Status of Constants in DD:}

\begin{itemize}
    \item DD claims all constants are \textbf{computable in principle}
    \item Currently: qualitative relations and constraints
    \item Key prediction: $\Lambda$ evolves (testable via DESI)
    \item Open problem: derive specific numerical values
\end{itemize}

The constants are not free parameters. They are fixed by self-consistency. We just haven't computed them all yet.
}}
\end{center}
