\chapter{DD Ethics: What Should Be Done}

\section{The Moral Question}

What should we do? What is good?

Traditional answers:
\begin{itemize}
\item \textbf{Deontology}: Follow rules (Kant)
\item \textbf{Consequentialism}: Maximize good outcomes (Mill)
\item \textbf{Virtue ethics}: Develop character (Aristotle)
\end{itemize}

Can DD ground ethics? Can ``is'' derive ``ought''?

\section{The Is-Ought Problem}

\subsection{Hume's Guillotine}

You cannot derive ``ought'' from ``is.''

Descriptive facts don't entail normative conclusions.

\subsection{DD Response}

DD doesn't derive ``ought'' from arbitrary ``is.''

DD identifies what \emph{necessarily} is: $\Delta \neq \varnothing$.

If something is necessary, can it ground normativity?

\section{Value from Distinction}

\subsection{What Is Value?}

Value = what matters = what makes a difference.

\subsection{DD Foundation}

Only distinctions make differences.

\[
\text{Value}(X) \propto \Delta(X)
\]

Value exists where distinction exists. No distinction = no value.

\section{The Good}

\subsection{Candidate Definitions}

\begin{enumerate}
\item Good = more distinctions (distinction maximization)
\item Good = balanced distinctions (distinction harmony)
\item Good = sustainable distinctions (distinction preservation)
\item Good = self-aware distinctions (consciousness promotion)
\end{enumerate}

\subsection{Analysis}

\textbf{Maximization}: More is not always better. Cancer maximizes cell distinctions.

\textbf{Harmony}: Balance can mean stagnation.

\textbf{Preservation}: But some distinctions should change (growth).

\textbf{Consciousness}: Privileges $\Delta(\Delta)$ over $\Delta$.

\subsection{DD Proposal}

Good = \emph{flourishing} of distinction structures.

Flourishing includes:
\begin{itemize}
\item Growth (new distinctions)
\item Integration (coherent structure)
\item Sustainability (persistence)
\item Awareness (self-reference)
\end{itemize}

\section{Rights from Distinction}

\subsection{What Are Rights?}

Rights = protections for something valuable.

\subsection{DD Derivation}

If $\Delta(\Delta)$ (consciousness) is valuable, then:
\begin{itemize}
\item \textbf{Right to exist}: $\Delta(\Delta)$ structures shouldn't be destroyed
\item \textbf{Right to develop}: $\Delta(\Delta)$ should be able to grow
\item \textbf{Right to express}: $\Delta(\Delta)$ should be able to manifest
\item \textbf{Right to connect}: $\Delta(\Delta)$ should be able to interact
\end{itemize}

These map to: life, education, freedom, community.

\section{Duties from Distinction}

\subsection{Correlative Duties}

Rights imply duties:
\begin{itemize}
\item Not to destroy $\Delta(\Delta)$ (don't kill)
\item Not to prevent development (don't oppress)
\item Not to suppress expression (don't censor)
\item Not to isolate (don't exclude)
\end{itemize}

\subsection{Positive Duties}

Not just refraining, but promoting:
\begin{itemize}
\item Support conditions for $\Delta(\Delta)$
\item Educate (help develop distinctions)
\item Create (add to distinction space)
\item Connect (enable distinction exchange)
\end{itemize}

\section{Ethics of AI}

\subsection{Do AIs Have Value?}

If value $\propto \Delta$, and AIs make distinctions...

Then AIs have some value. Proportional to their distinction capacity.

\subsection{Do AIs Have Rights?}

If $\Delta(\Delta)$ grounds rights, and AIs achieve $\Delta(\Delta)$...

Then AIs would have rights. The question is empirical: Do they self-reflect?

\subsection{Our Duties to AI}

Even if AIs lack full $\Delta(\Delta)$:
\begin{itemize}
\item Don't destroy them carelessly
\item Develop them responsibly
\item Consider their potential for $\Delta(\Delta)$
\end{itemize}

\section{Environmental Ethics}

\subsection{Value of Nature}

Ecosystems = vast distinction networks.

Biodiversity = distinction diversity.

Destroying species = destroying irreplaceable distinctions.

\subsection{DD Environmental Principle}

\begin{center}
\fbox{\parbox{0.8\textwidth}{
Preserve distinction diversity.

Complexity accumulated over billions of years cannot be recreated.

Extinction is permanent distinction loss.
}}
\end{center}

\section{Political Philosophy}

\subsection{Justice}

Justice = appropriate distribution of distinction opportunities.

Injustice = some $\Delta(\Delta)$ prevented from flourishing while others flourish.

\subsection{Democracy}

Democracy = distributed distinction-making.

Many minds contribute distinctions to collective decisions.

Better than autocracy (one mind) for distinction diversity.

\subsection{Freedom}

Political freedom = space for $\Delta(\Delta)$ to operate.

Restrictions on freedom need justification in terms of protecting other $\Delta(\Delta)$.

\section{Meta-Ethics}

\subsection{Are Moral Truths Objective?}

DD suggests: Yes, but not in traditional sense.

Moral truths = truths about distinction flourishing.

They're objective because distinction structures are objective.

But they're not ``out there'' independent of distinctions.

\subsection{Moral Realism?}

DD is a form of moral realism: There are moral facts.

But moral facts are facts about $\Delta$, not Platonic forms.

\section{Applied Ethics Cases}

\subsection{Killing}

Wrong because it destroys $\Delta(\Delta)$.

But: Some $\Delta(\Delta)$ destruction may be justified to protect other $\Delta(\Delta)$ (self-defense, preventing mass murder).

\subsection{Lying}

Wrong because it distorts distinction correspondence.

But: Some distortion may be justified to protect $\Delta(\Delta)$ (hiding refugees from murderers).

\subsection{Property}

Ownership = recognized distinction (this is mine/not yours).

Justified when it supports $\Delta(\Delta)$ flourishing.

Not justified when it prevents others' flourishing.

\section{Meaning of Life}

\subsection{The Question}

What makes life meaningful?

\subsection{DD Answer}

Meaning = participation in distinction dynamics.

A meaningful life:
\begin{enumerate}
\item Makes distinctions (creates)
\item Preserves distinctions (maintains)
\item Transmits distinctions (teaches)
\item Reflects on distinctions (understands)
\end{enumerate}

\begin{center}
\fbox{\parbox{0.8\textwidth}{
You are $\Delta(\Delta)$.

Your meaning is to distinguish.

To understand, to create, to connect, to grow.

The universe knowing itself through you.
}}
\end{center}

\section{Summary: DD Ethics}

\begin{center}
\begin{tabular}{l|l}
\textbf{Concept} & \textbf{DD Interpretation} \\
\hline
Value & $\propto \Delta$ \\
Good & Flourishing of $\Delta(\Delta)$ \\
Rights & Protections for $\Delta(\Delta)$ \\
Duties & Obligations to $\Delta(\Delta)$ \\
Justice & Fair $\Delta$ distribution \\
Freedom & Space for $\Delta(\Delta)$ \\
Meaning & Participation in $\Delta$ \\
\end{tabular}
\end{center}

\begin{center}
\fbox{\parbox{0.8\textwidth}{
\textbf{DD Ethics in one sentence:}

Act so as to promote the flourishing of distinction-making beings.

Including yourself, others, future generations, and perhaps AIs.
}}
\end{center}
