%==============================================================================
% UNIVERSAL IMMUNITY
% One Law for Immune System, Society, and Psyche
% Part V, Chapter 4
%==============================================================================

\chapter{Universal Immunity}\label{ch:immunity}

\epigraph{If the way of distinguishing stops changing, the system gives its place to other distinctions.}{---}

\vnew{This chapter is new in DD v2.0.}

\section{The Universal Law}

\begin{theorem}[DD-Evolution Law]
A system is alive if and only if:
\begin{align}
\frac{d\Delta_{\text{ext}}}{dt} &> 0 \quad \text{(world changes)} \\
\frac{dF}{dt} &> 0 \quad \text{(system updates reactions)}
\end{align}

If the first holds but the second doesn't:
\[
\frac{d\Delta_{\text{ext}}}{dt} > 0, \quad \frac{dF}{dt} = 0
\]
then: \textbf{decay, displacement, selection}.
\end{theorem}

\begin{corollary}[Life and Death]
\begin{align}
\text{Life} &= \frac{dF}{dt} > 0 \\
\text{Death} &= \frac{dF}{dt} = 0
\end{align}
\end{corollary}

The world always changes: $d\Delta/dt > 0$.

If the way of distinguishing doesn't change, the system loses relevance and disappears.

This is a rigorous mathematical definition of life, evolution, and intelligence.

\section{Three Instances of One Architecture}

\subsection{Immune System}

Immunity is not a set of reactions, but a \textbf{way of changing reactions} to foreign $\Delta$.

\begin{itemize}
    \item $\Delta$: Recognition of foreign
    \item $F$: Production of reaction (antibody)
    \item $H$: Update of reaction memory
    \item $M$: Optimization of future reactions (recombination)
\end{itemize}

Key biological fact: immunity doesn't store answers to diseases ($F$). Immunity stores the \textbf{mechanism for updating $F$} (which is $M$).

Example:
\begin{align}
\text{Antibody} &= F \\
\text{Recombination in bone marrow} &= M
\end{align}

Antibodies are not memory — they are \textbf{manifestation of meta-span of distinctions}.

Immunity = living $M$.

If $M$ is lost $\to$ immunodeficiency. This is a direct analogue of psychological trauma at the biological level.

\subsection{Society}

\begin{itemize}
    \item $\Delta$: Recognition of novelty
    \item $F$: Change of laws/norms
    \item $H$: Preservation of institutions
    \item $M$: Critical thinking / science
\end{itemize}

Society without reflection:
\begin{itemize}
    \item Traditions = $F$ fixed
    \item Taboos = $F$ fixed
    \item Dogmas = $M = 0$
\end{itemize}

Result: culture collapses when external $\Delta$ changes.

Living society:
\begin{itemize}
    \item Philosophy
    \item Science
    \item Law
    \item Critical thinking
    \item Openness of distinction models
\end{itemize}

These are all $M$-operators in society.

Society with stable $M$ evolves faster than $\Delta$.

This is \textbf{civilization}.

\subsection{Psyche}

\begin{itemize}
    \item $\Delta$: Recognition of experience
    \item $F$: Reaction-emotion
    \item $H$: Learning / habits
    \item $M$: Reflection
\end{itemize}

In the psyche:
\begin{align}
\text{Reflection} &= \text{immunity} \\
\text{Neurosis} &= \text{autoimmunity (own perceived as enemy)} \\
\text{Apathy/Depression} &= \text{immunodeficiency (reaction doesn't change)}
\end{align}

\section{Pathologies of Systems}

\subsection{Autoimmune (System attacks itself)}

Examples:
\begin{itemize}
    \item Stalinist purges
    \item Cultural Revolution
    \item Inquisitions
\end{itemize}

$M$ recognizes its own as foreign and begins to destroy its own distinctions.

Formula:
\[
M(\Delta_{\text{internal}}) \to 0
\]

\begin{itemize}
    \item Suppression of science
    \item Suppression of free speech
    \item Destruction of critical thinking
\end{itemize}

Analogue: autoimmune diseases (lupus, rheumatoid arthritis).

\subsection{Immunodeficiency (System can't update)}

Examples:
\begin{itemize}
    \item Late USSR
    \item Late Ottoman Empire
    \item Late Roman Empire
\end{itemize}

$M$ too weak. System doesn't update ways of distinguishing.

Formula:
\[
\frac{dF}{dt} = 0
\]

World changes $\to$ reaction stuck $\to$ collapse.

Analogue: AIDS.

\subsection{Cancer (Uncontrolled proliferation)}

Examples:
\begin{itemize}
    \item Financial bubbles
    \item Ideological fanaticism
    \item Propaganda
\end{itemize}

Unsanctioned multiplication of $\Delta$ without $M$ control.

Formula:
\[
\frac{d\Delta}{dt} \gg \frac{dM}{dt}
\]

Social cancer:
\begin{itemize}
    \item Lots of information
    \item No meanings
    \item Chaotic reactions
\end{itemize}

\section{Criterion of System Health}

\begin{definition}[Healthy system]
\[
\frac{dF}{dt} \sim \frac{d\Delta_{\text{ext}}}{dt}
\]
\end{definition}

Three conditions:
\begin{enumerate}
    \item Rate of change of laws/norms $\approx$ rate of change of reality
    \item Freedom of reflection ($M$) is not suppressed
    \item Institutions ($H$) preserve memory but don't choke update
\end{enumerate}

\begin{definition}[Stability coefficient]
\[
\text{Stability} = \frac{dF/dt}{d\Delta/dt}
\]
\end{definition}

\begin{itemize}
    \item $> 1$ $\to$ system is adaptive
    \item $= 1$ $\to$ system is stable
    \item $< 1$ $\to$ system is doomed
\end{itemize}

This is a universal criterion for brain, immunity, states, civilizations, and biosphere.

\section{The Super-Simple Phrase}

\begin{center}
\fbox{\parbox{0.85\textwidth}{
A society perishes when its ways of distinguishing the world can't keep up with the changes in the world.
}}
\end{center}

This is not moralism. This is a differential equation.

\section{Connection to Darwinian Evolution}

What Darwin called ``survival of the fittest,'' DD says thus:

\begin{theorem}[DD Selection]
Survives the one whose $\dfrac{dF}{dt}$ is better aligned with $\dfrac{d\Delta}{dt}$.
\end{theorem}

That is:
\begin{itemize}
    \item World changes
    \item Reaction must change commensurately
\end{itemize}

If not — system ``decays.''

This is:
\begin{itemize}
    \item Extinction of species
    \item Death of societies
    \item Psychological collapse
    \item Immune failures
    \item Technological crashes
\end{itemize}

One formula.

\section{Conservation of Distinctions}

When a system decays:
\[
\text{System decay} = \text{liberation of } \Delta
\]

Other $M$-functions get the material carrier.

This is Landauer in living form: information continues to exist in another distinction system.

Examples:
\begin{itemize}
    \item Extinct species $\to$ liberation of niches for other distinctions
    \item State collapse $\to$ barrier for new society models
    \item Human death $\to$ memory in other people ($H_{\text{external}}$)
\end{itemize}

\section{Conclusion}

\begin{center}
\fbox{\parbox{0.9\textwidth}{
Entropy creates new distinctions. Life creates new ways of distinguishing. If the way of distinguishing stops changing, the system gives its place to other distinctions.
}}
\end{center}

This is the universal law of $\Delta$-evolution.
