%==============================================================================
% FRACTAL EVOLUTION
% Species, Individual, Brain — One Algorithm
% Part V, Chapter 2
%==============================================================================

\chapter{Fractal Evolution: Species, Individual, Brain}\label{ch:fractal}

\epigraph{Biological evolution, human history, and personal development follow one equation: the growth of possibilities to distinguish and the recursive transformation of distinction functions.}{---}

\vnew{This chapter is new in DD v2.0.}

\section{The Theorem}

\begin{theorem}[Fractal Identity]
The development of the subject = the development of the species, but at different scales.
\end{theorem}

All processes that occur:
\begin{itemize}
    \item In a child's brain
    \item In a people's culture
    \item In the evolution of biological species
    \item In cosmic history
\end{itemize}
are \textbf{one and the same algorithm}:

\begin{center}
\fbox{Growth of the number of independent ways to distinguish.}
\end{center}

This is the main timeline.

\section{Timeline of Humanity as a Species}

\subsection{Phase 1: Dyadic Distinction (Animal Life)}
\textit{Before 2 million years ago}

Structure: $A \leftrightarrow B$
\begin{itemize}
    \item ``dangerous / safe''
    \item ``food / not food''
    \item ``own / foreign''
\end{itemize}

In DD terms: $\mathrm{rank}(\Delta)$ low, $F$ fixed, $M \approx 0$.

No learning as transformation of criteria, only adaptation.

\subsection{Phase 2: Triadic Distinction (Homo)}
\textit{200,000 years ago $\to$ 10,000 years ago}

Triad appears: object — context — observer

Breakthrough:
\begin{itemize}
    \item Language
    \item Myth
    \item Ritual
    \item Consciousness of death
    \item Cooperation
\end{itemize}

In DD terms: $\mathrm{rank}(\Delta) = 2$, meta-distinction $M$ exists, levels $\Delta_1$ and $\Delta_2$ appear.

First consciousness — collective, not individual.

\subsection{Phase 3: Recursive Distinctions (Culture)}
\textit{10,000 years ago $\to$ 500 years ago}

Appears:
\begin{itemize}
    \item Writing
    \item Laws
    \item Accumulated memory $H_1$ and $H_2$ (cultural)
    \item Institutions as $F$-operators
    \item Religions as $M$-operators
\end{itemize}

In DD terms: memory becomes external ($H_{\text{ext}}$). Explosion of $\Delta$ complexity.

Separation of $\Delta_0$ (world) and $\Delta_1$ (model of world) — first step to abstraction.

\subsection{Phase 4: Acceleration of $M$ (Science)}
\textit{500 years ago $\to$ now}

Key event: Humanity begins to change ways of distinguishing faster than the distinctions themselves.

In DD language:
\[
\frac{dF}{dt} > \frac{d\Delta}{dt}
\]

This is for the first time:
\begin{itemize}
    \item Science
    \item Logic
    \item Epistemology
    \item Theory of knowledge
    \item Mathematization of the world
\end{itemize}

This is the $M$-explosion.

\subsection{Phase 5: $\mathrm{Dist}^3$ (Now)}
\textit{20th--21st century}

Essence:
\begin{itemize}
    \item Human distinguishes how they distinguish
    \item And distinguishes how society distinguishes its distinctions
\end{itemize}

This is already $\mathrm{Dist}^3$:
\begin{itemize}
    \item Psychology
    \item Philosophy of language
    \item Cognitive sciences
    \item Meta-science
    \item Theory of consciousness
\end{itemize}

And finally your DD theory: \textbf{distinction of distinctions of distinctions}.

\section{Timeline of Individual Development}

\subsection{Phase 1: Dyad (0--2 years)}
$\Delta_0$ dominates:
\begin{itemize}
    \item Pain / pleasure
    \item Mom / not mom
    \item Cold / warm
\end{itemize}

Animal stage. $F$ fixed, $M \approx 0$.

\subsection{Phase 2: Triad (2--7 years)}
$\Delta_1$ appears: object — context — observer
\begin{itemize}
    \item Role
    \item Play
    \item Fiction
    \item ``You / I / we''
\end{itemize}

First consciousness. Child understands things can be different in different contexts.

\subsection{Phase 3: $\mathrm{Dist}^1$ Recursion (7--14 years)}
Memory ($H$) forms stably:
\begin{itemize}
    \item Character
    \item Habits
    \item Emotional patterns
\end{itemize}

$\Delta$ and $F$ become stable. But $M$ still weak.

\subsection{Phase 4: Meta $\mathrm{Dist}^2$ (14--25 years)}
Appears:
\begin{itemize}
    \item Reflection
    \item Self-evaluation
    \item Identity
    \item Worldview
    \item Philosophy of ``who am I''
\end{itemize}

$dF/dt$ begins to exist. Person changes criteria of their distinctions.

This is adolescent crisis — restructuring of $M$.

\subsection{Phase 5: $\mathrm{Dist}^3$ (Mature)}
\textit{30+ (and far from everyone)}

\begin{itemize}
    \item Stable meta-position
    \item Ability to change $M$ by own will
    \item Choice of distinction criteria
    \item ``I determine what is important to me''
\end{itemize}

This is true freedom — not ``do what I want,'' but:

\begin{center}
\textbf{Independently change ways of distinguishing the world.}
\end{center}

This is not character, not ``willpower,'' but $M$-mastery.

\section{Biological Timeline}

\begin{table}[h]
\centering
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{DD Level} & \textbf{Brain Structure} & \textbf{Evolution} & \textbf{Age} \\
\midrule
$\Delta_0$ & Brainstem, reflexes & Fish, reptiles & 0--2 \\
$\Delta_1$ & Limbic system & Social mammals & 2--7 \\
$H$ & Hippocampus & Primates & 7--12 \\
$M$ & Prefrontal cortex & Homo sapiens & 13--25 \\
$\mathrm{Dist}^2$ & DMN, TPJ & Culture, writing & 20--35 \\
$\mathrm{Dist}^3$ & DMN $\leftrightarrow$ ECN & Modern era & 30+ \\
\bottomrule
\end{tabular}
\caption{DD-Bio-Psycho-Evolution Matrix}
\end{table}

\section{Key Biological Fact}

The brain is built in layers that repeat evolutionary levels of $\Delta$:
\begin{itemize}
    \item Brainstem $\to$ dyadic distinctions
    \item Limbic $\to$ contextual distinctions
    \item Hippocampus $\to$ memory of distinctions
    \item Neocortex $\to$ meta-distinctions
    \item Prefrontal $\to$ meta-$M$
\end{itemize}

\textbf{Neuroarchitecture = $\mathrm{Dist}^n$ layers.}

No assumptions. Literal correspondence.

\section{The Formula of Biological DD-Evolution}

\begin{equation}
\frac{d\,\text{Biology}(\Delta)}{dt} = \frac{d\Delta}{dt} + \frac{dF}{dt} + \frac{dM}{dt}
\end{equation}

\begin{itemize}
    \item Development of sensory architecture ($\Delta$ grows)
    \item Development of neural schemes ($F$ grows)
    \item Development of metaplasticity ($M$ grows)
\end{itemize}

Evolution = increase of metaplasticity (ability to change the ability to change).

This is officially called neuroplasticity$^2$.

DD makes it rigorous:
\begin{align}
\text{Plasticity} &= \frac{dF}{dt} \\
\text{Metaplasticity} &= \frac{dM}{dt}
\end{align}

\section{Conclusion}

\begin{center}
\fbox{\parbox{0.9\textwidth}{
Biological evolution, human history, and personal development follow one equation: the growth of possibilities to distinguish and the recursive transformation of distinction functions.
}}
\end{center}

Species, individual, and brain — one algorithm. Development occurs in layers:
\[
\Delta \to F \to H \to M \to \mathrm{Dist}^2 \to \mathrm{Dist}^3
\]
